Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-09-16 writeup/pnas-word-order-si.tex
SI

\item Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}} is formalized using \emph{nmod} which covers not only genitives, but also all other noun-modifying NPs and PPs.
	The evaluation of extracted grammars against WALS (Table~\ref{tab:grammars-wals}) shows that, among the 37 languages where WALS has an entry, the dominant direction of \emph{nmod} agrees with that of genitives, with two exceptions (Danish and Swedish).

@@

In Figure~\ref{fig:pareto-per-lang}, we show the predictability-parseability planes for every one the 51 languages, together with Pareto frontiers estimated from optimized grammars.
We also provide the raw numerical values, before $z$-scoring, in Figure~\ref{fig:pareto-per-lang-raw}.

@@

>>>>

\item Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}} is formalized using \emph{nmod} which covers not only genitives, but also all other noun-modifying NPs and PPs.
	The evaluation of extracted grammars against WALS (Table~\ref{tab:grammars-wals}) shows that, among the 37 languages where WALS has an entry, the dominant direction of \emph{nmod} agrees with that of genitives, with two exceptions (Danish and Swedish\footnote{Danish and Swedish have genitives preceding the head marked with -\emph{s} similar to English (reflected in the WALS entry), while noun-modifying PPs, including phrases similar to English \emph{of} phrases, follow the head. In these two languages, the order of adnominal PPs, agreeing with the more frequent order of \emph{nmod} relations, agrees with the verb-object relation, whereas prenominal -s genitives show the opposite ordering.}.

@@

\subsection{Per-Language Results}
In Figure~\ref{fig:pareto-per-lang}, we show the predictability-parseability planes for every one the 51 languages, together with Pareto frontiers estimated from optimized grammars.
We also provide the raw numerical values, before $z$-scoring, in Figure~\ref{fig:pareto-per-lang-raw}.

@@

\subsection{Analysis controlling for Families}

The UD treebanks overrepresent certain language families.
Does this impact the results of Study 1?
We estimate the overall degree of optimization of languages for efficiency, controlling for differences between families, we constructed a Bayesian logistic mixed-effects model estimating, for each language $L$ among the 51 UD languages, the rate $q_L \in [0,1]$ of random baseline grammars that have lower efficiency (parseability, predictability) than the real grammar.
We entered languages and language families as random effects:
\begin{equation}\label{eq:mixed-effects-study1}
logit(q_{L}) = \beta + u_{L} + v_{f_L}
\end{equation}
where $f_L$ is the language family of $L$.

Results for the posterior of $\beta$ are shown in Table~\ref{tab:study1}.
Across language families, real grammars tend to be more efficient than almost all baseline grammars.

\begin{table}
\small{
\begin{center}
\begin{tabular}{|l||l|lll|llll|ll|llllll}
\hline
	& Mean $\beta$ & SD & Lower 95\% CrI & Upper 95\% CrI \\
\hline\hline
	Efficiency ($\lambda=0.9$) &	-5.88   &   1.08  &  -8.28  &  -3.97  \\
	Surp & -3.48   &   0.88  &  -5.42  &  -1.85   \\
	Pars & -5.55   &   1.08  &  -7.80  &  -3.67    \\
\hline
\end{tabular}
\end{center}
}
	\caption{TODO}\label{tab:study1}
\end{table}
% paretoByFamily.R

\subsection{Quantifying Pareto-Optimality}

\paragraph{Inferring $\lambda$}
We choose $\lambda$ to minimize the number of baselines that have better values of $R_\lambda := R_{Pars} + \lambda R_{Pred}$ than the real language:
\begin{equation}
	O := \min_\lambda \# \{b \in B: R_{eff}(b) > R_{eff}(r)\}
\end{equation}
where $r$ is the real language, and $B$ is the set of baselines.
Naively choosing $\lambda$ to minimize this quantity from the observed $50$ baseline samples will bias $O$ towards overestimating the optimality of the real language.

TODO maybe add a figure explaining the Bayesian story a bit more -- how we're using a continuous distribution to smooth out the observed samples?

We eliminate this anti-conservative bias by conducting joint Bayesian inference of $\lambda$ of $O$, using the Dirichlet Gaussian Process, a Bayesian version of Kernel Density Smoothing.
The 50 samples represent an underlying baseline distribution in the predictability-parseability plane.
We approximate this underlying distribution as a continuous distribution from the Dirichlet Gaussian Process, and infer its parameters using Bayesian inference.
We use this method to obtain a Bayesian estimate of the distribution of baseline distributions in the Surprisal-Parseability plane.
The Dirichlet Gaussian Process (CITE) defines a prior over possible baseline distributions.
Under this prior, each possible distribution is parameterized by a number $n$ of components, and $n$ Gaussians $\mu_n, \sigma^2_n$.
Due to the flexible number of components, this nonparametric prior can accomodate a wide range of continuous probability distributions, including multimodal ones.
We use the $50$ observed baseline samples to infer a posterior distribution over such smooth distributions.
We then use MCMC to obtain 1,000 samples from the posterior, based on these samples.
We used the MCMC implementation provided by the \texttt{R} package \texttt{dirichletprocess}~\citep{ross2018dirichletprocess}, discarding the first 1,000 samples as warmup samples, and obtaining another 1,000 samples.

For every sampled distribution -- consisting of a component number $n$ and $n$ Gaussians, we then computed 
\begin{equation}
	O := \min_\lambda P(R_{eff}(b) > R_{eff}(r))\}
\end{equation}
where we considered $\lambda \in (0,1)$ in steps of $0.5$.

By taking the values $O, \lambda$ for the 1,000 posterior samples, we obtained an estimate of the posterior of $O$ and $\lambda$.

\paragraph{Analysis across Values of $\lambda$}

We quantify the Pareto-optimality of the language $r$ as the fraction of baselines that are better than $r$ on \emph{both} parseability and predictability:
\begin{equation}
	PO := \frac{\# \{b \in B: R_{Pars}(b) > R_{eff}(r) \text{ and }  R_{Pred}(b) > R_{Pred}(r) \}}{\#B}
\end{equation}
This quantifies how many hypothetical grammars are more efficient than $r$ \emph{irrespective} of $\lambda$.

Again, we use the Dirichlet Gaussian Process to obtain an estimate of the posterior of $PO$.
We used the same procedure to also estimate the curve connecting points $x$ in the predictability-parseability that have the same value $PO(x) = PO(r)$ -- this curve represents which baselines 
Again, due to the smoothing implied by the Dirichlet Gaussian Process model, this curve connects not observed baseline samples, but hypothetical other baselines in the predictability-parseability plane.

- identify lambda and quantify what fraction of the baselines are less efficient at this lambda (this is essentially the same as what Zaslavsky et al 2018 do)

- also to make sure this isn't just because lambda can be chosen, quantify what fraction of baselines Pareto-dominate the same fraction
